Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model

### 2023-01-06 src/docs_paper/appendix.Rmd
Work on moving model into main paper and improving appendix description

Consider some country, and let $x$ refer to a district located within the Spectrum region $R_x$.
We restrict our attention to the time period $t = T_1$ of the most recent national household survey with HIV testing.
Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$ be five-year age bands.
As short-hand, we write $a = l$ to refer to the age band with lower bound $l$.
We index the following quantities by district, sex, age-band: population size $N_{x, s, a}$, HIV prevalence $\rho_{x, s, a}$, ART coverage $\alpha_{x, s, a}$ and annual HIV incidence rate $\lambda_{x, s, a}$.
Our model operates at this level of granularity, although sometimes the data observed is at an aggregate level.
We use $\{\cdot\}$ to refer to a generic aggregate set over which an observation is made, such as $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for adults.

@@

Consider some generic random effect $u$.
We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with marginal standard deviation $\sigma > 0$.
For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising] with marginal standard deviation $\sigma > 0$ we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component with proportion $\phi \in (0, 1)$ and spatially unstructured IID component with proportion $1 - \phi$
We follow recommendations of @freni2018note on scaling, disconnected adjacency graph components, and islands.
The first order auto-regressive model is specified by $u \sim \text{AR}1(\sigma, \phi)$.

@@

We model HIV prevalence by
$$
\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a} 
$$
where $\beta^\rho_0$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}}$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}}$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
We use the prior distributions $\mathcal{N}(0, 5)$ for the fixed effects $\beta^\rho_0$ and $\beta_{s}^{\rho, s = \text{M}}$, $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms, $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters, and $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters.

@@

Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person of ART residing in district $x$ recieves ART in district $x'$.
We assume that $\gamma_{x, x'} = 0$ for $x \notin \{x, \text{ne}(x)\}$ such that individuals seek treatment only in their residing district and its neighbours $\text{ne}(x) = \{x': x' \sim x\}$, where $\sim$ is an adjacency relation and $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x'} = 1$.

>>>>

Consider a given country in sub-Saharan Africa.
Let $x$ refer to a district located within the Spectrum region $R_x$.
We restrict our attention to the time period $t = T_1$ of the most recent national household survey with HIV testing.
Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$ be five-year age bands.
As short-hand, we write $a = l$ to refer to the age band with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
We index the following quantities by district, sex and age-band: population size $N_{x, s, a}$, HIV prevalence $\rho_{x, s, a}$, ART coverage $\alpha_{x, s, a}$, annual HIV incidence rate $\lambda_{x, s, a}$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a}$.
Sometimes data are observed at an aggregate level, rather then the level of granularity at which the model operates.
In this instance, we use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for adults.

@@

Consider a generic random effect $u$.
We use use structured random effects to share information across units assessed as being similar, such as districts next to each other or adjacent age-bands.
We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
$$
u_i \, | \, u_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
$$
where $u_{-i}$ is $u$ with the $i$th unit removed, $j \sim i$ if the units are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands regarding the ICAR model.
For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising] we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v^\star$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w^\star$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
$$
u = \sigma \left( \sqrt\phi \cdot v^\star + \sqrt{1 - \phi} \cdot w^\star \right).
$$
We specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
\begin{align*}
u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
u_i &= \rho u_{i - 1} + \epsilon_t, \quad i = 2, \ldots
\end{align*}
where $\epsilon_i \sim \mathcal{N}(0, 1)$ is Gaussian white noise, $|\rho| < 1$ is the lag-one correlation parameter.

@@

We model HIV prevalence on the logit scale using
$$
\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a} 
$$
where $\beta^\rho_0$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}}$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}} \sim \text{UNKNOWN}$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
We use the prior distributions $\mathcal{N}(0, 5)$ for the fixed effects $\beta^\rho_0$ and $\beta_{s}^{\rho, s = \text{M}}$, $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms, $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters, and $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters.

@@

Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ recieves ART in district $x'$.
We assume that $\gamma_{x, x'} = 0$ for $x \notin \{x, \text{ne}(x)\}$ such that individuals seek treatment only in their residing district and its neighbours $\text{ne}(x) = \{x': x' \sim x\}$, where $\sim$ is an adjacency relation and $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x'} = 1$.

### 2023-01-06 src/docs_paper/paper.Rmd
Work on moving model into main paper and improving appendix description

For each country, the full model is defined over three time points: $T_1$, the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
We consider a simplified version of the full model which is defined only at $T_1$ and omits temporal projection.
In this section, we provide an overview of the modified model, highlighting the aspects which make it a challenge for existing inferential approaches.
Full mathematical details of this modified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.

## Process model

Let $i$ index the coarsest space-age-sex division included in the model.
Let $N_i$ be the population size and consider the following indicators: HIV prevalence $\rho_i$, annual HIV incidence rate $\lambda_i$ and ART coverage $\alpha_i$.

## Likelihood

To write.

>>>>

Let $x$ index district, $a$ index five-year age band, $s$ index sex and $t$ index time.
For each country, the full model is defined over three time points: $T_1$, the time of the most recent household survey with HIV testing; $T_2$, the current time period; and $T_3$, a short term projection period.
We consider a simplified version defined only at $T_1$, omiting all temporal projection.
In this section, we provide an overview of the modified model, highlighting the aspects which make it a challenge for existing inferential approaches.
Full mathematical details of this modified model, as well as a \textsf{C++} template for the log-posterior, are provided in the appendix.

## Household survey component

Consider a household survey occurring at $T_1$, and let $i$ index the coarsest district-age-sex division included in the model.
The data we observe may be aggregated over indices $i$, so we let $\mathcal{I}$ be a set of $i$ for which observations are reported.
Let $N_i$ be the population size, and consider the following three indicators: HIV prevalence $\rho_i \in [0, 1]$, ART coverage $\alpha_i \in [0, 1]$, and annual HIV incidence rate $\lambda_i > 0$.
We specify independent mixed effects models for HIV prevalence and ART coverage in the general population on the logit scale such that
\begin{align*}
\text{logit}(\rho_i) &= \eta^\rho_i, \\
\text{logit}(\alpha_i) &= \eta^\alpha_i,
\end{align*}
for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
For the HIV incidence rate we use a mixed effects model on the log scale
$$
\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}),
$$
where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$.

Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
For each set of observed strata indices $\mathcal{I}$, we calculate the weighted observations $\hat \theta_\mathcal{I}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes 
$$
M^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
$$
where $j$ index individuals in all strata $i \in \mathcal{I}$, with corresponding survey weights $w_j$.
The observed number of indicator cases is then
$$
Y^{\hat \theta}_{\mathcal{I}} = M^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}.
$$
For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood
$$
Y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(M^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}),
$$
where $\theta_{\mathcal{I}}$ are the following weighted aggregates
\begin{align*}
\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \\
\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \\
\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
\end{align*}
We link the proportion recently infected among HIV positive persons $\kappa_i$ to HIV incidence $\lambda_i$ by
$$
\kappa_i = 1- \exp \left( - \lambda \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
$$
where $\Omega_T$ is the mean duration of recent infection and $\beta_T$ is the false recent ratio.

## ANC testing component

We model HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women as being offset on the logit scale from the general population indicator as follows
\begin{align*}
\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + \tilde{\eta}^{\rho^\text{ANC}}_i, \\
\text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \tilde{\eta}^{\rho^\text{ANC}}_i.
\end{align*}
Process section to be written.

Likelihood section to be written.

## ART attendance component

Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ recieves ART in district $x'$.
Process section to be written.

Let $\dot A_i$ be the number of people receiving ART.
Likelihood section to be written.
